---
title: "Exploration of biomass modelling options"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: resources/Rmd-html-style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
options(tinytex.engine = 'xelatex')
knitr::read_chunk('../scripts/biomass_tests.R')
```

# Synopsis

The intention of this document and associated R script is to motivate
a discussion about options for modelling fish biomass.  The AIMS
Long-term monitoring program collects fish Abundance and length data
annually from a fixed set of transects (within sites within reefs).
Whilst, we can (and do) model fish Abundance, there is a strong desire
to express fish trends in terms of biomass.

Biomass is not directly measured.  Rather, it is calculated from
Abundances and lengths (the exact calculations vary from species to
species).  Since biomass is calculated from both Abundance and Length,
it has a mixture of count and length properties.  Specifically, it is
comprises of continous positive values as well as substantial zero
values (likely to be false positives associated with imperfect
detection).

This document will explore options for modelling such data by first
attempting to simulate relatively complex and messy data.  Thereafter,
there will be attempts to fit models to explore temporal trends in
both Abundance and Biomass.


# Package setup

```{r libraries, warnings=FALSE, message=FALSE, results='hide'}
```

```{r dirs, results='markdown', eval=TRUE, echo=FALSE}=
```


# Data simulation

The observed data will comprise of:

- a hierarchical sampling design in which there are five transects
  nested within sites each of three sites which are in turn nested
  within each of reefs 10.
- along each transect, whenever a group of fish are detected (= 1 or
  more individuals of the same species), the species and number of
  individuals along with the approximate average body length are
  recorded
  
 | Reef | Site | Transect | Species | Year | Abundance | Length |
 |------|------|----------|---------|------|-----------|--------|
 |      |      |          |         |      |           |        |
 
**Note**, since fish are recorded in groups, a single species in a
transect might be represented by multiple rows (one for each
encountered groups.)

<details><summary>Details of the data synthesis</summary>

1. Lets start by defining the spatial design (6 Reefs, each with 3
   Sites, each with 5 transects)

```{r synthetic-data-1, results='markdown', eval=TRUE}
```
2. Now for each of eight fish species, we will define separate splines
   that represent the different temporal trends of each species

```{r synthetic-data-2, results='markdown', eval=TRUE, figwidth=6, fig.height=3}
```

3. propogate these splines (with 10 evenly spaced time units) in the
   data to create a wide data set before pivoting into a long data set

```{r synthetic-data-3, results='markdown', eval=TRUE, fig.width=6, fig.height=3}
```

**Note** currently, the temporal patterns for each species are the same within each transect

4. currently, the synthetic data comprises the full temporal series
   for each species on each transect.  A fish sampling program will
   typically not encounter every species of fish within each transect
   on each sampling occassion - either because not every species
   occurs at each location or simply that each species are not
   encountered every time.  Furthermore, fish are often encounted in
   schools or groups.
   
   To mimic this, we will subsample the species within each
   transect/period (the number of groups of fish encountered will be
   drawn from a Poisson distribution.  Subsampling will be biased such
   that the probability of selecting (encountering) a species is
   proportional to its id number (e.g. Species 8 is much more common
   than Species 1).

```{r synthetic-data-4, results='markdown', eval=TRUE}
```

5. fish counts should follow a Poisson (or Negative Binomial)
   distribution.  This does however assume perfect detection.  When
   detectibility is not 100%, the observed data are likely to be zero
   inflated (whereby some absences really reflect an inability to
   detect the fish rather than true absenses of fish).
   
   To simulate this, we will generate zero-inflated Poisson data in
   which the probability of a false zero (failure to detect) is 0.2
   and the count process is governed by a linear predictor comprising:
   
   - the species specific temporal trend
   - the reef, site and transect effects
   - a species abundance multiplier
   
   At the same time, we also want to collect average fish lengths of
   the encountered groups of fish.  In doing so, we might assume that
   the more common fish would be smaller than the less common fish, so
   we will draw the fish lengths out of a gamma distribution whose
   mean is inversely proportional to the Species number and dispersion
   is equal to 1)

```{r synthetic-data-5, results='markdown', eval=TRUE, fig.width=7, fig.height=5}
```

</details>


The following interactive graphic provides a way to explore the
temporal trend of fish Abundance within each Transect conditional on
Species.  The dropdown selector provides a way to filter to specific
sites.

```{r interactive_fig1, results='markdown', eval=TRUE}
```


```{r interactive_fig2, results='markdown', eval=TRUE, fig.width=8, figh.height=10}
```



# Data processing

Although we could fit models to explore the temporal patterns in fish
Abundance, we might alternatively be more interested in exploring
patterns of biomass.  For many fish species, biomass can be calculated
from counts and lengths accouding to the formula

$$
biomass = a\times abundance\times length^b
$$

where $abundance$ and $length$ are the observed number of average
length of fish in a group and $a$ and $b$ arespecies specific
coefficients obtained from the fishbase database.

We are primarily interested in biomass as a sum across all encountered
species, yet we must fist calculated the biomass of fish in each
encountered group of fish using the above formula and species specific
coefficients.

```{r data-processing, results='markdown', eval=TRUE}
```

# Modelling options

## Abundance {.tabset .tabset-faded}

### ZINB (glmmTMB) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

```{r fitModel_zinb, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Explore diagnostics

```{r fitModel_zinb_diagnostics, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

**Conclusions:**

- no evidence of dispersion issues
- residual plot looks good
- no evidence of zero-inflation

#### Summarise model

```{r fitModel_zinb_summary, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

::: {width="100%"}

```{r fitModel_zinb_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

### ZINB (INLA) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

```{r fitModel_zinb_INLA, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Explore diagnostics

```{r fitModel_zinb_INLA_posteriors, results='markdown', eval=TRUE, cache=TRUE, fig.width=8, fig.height=5}
```

#### Summarise model

```{r fitModel_zinb_INLA_summary, results='markdown', eval=TRUE, cache=TRUE, fig.width=8, fig.height=5}
```

:::

::: {width="100%"}

```{r fitModel_zinb_INLA_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::


### ZINB (brms) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

```{r fitModel_zinb_brm, results='markdown', eval=FALSE, fig.width=8, fig.height=5, cache=FALSE}
```

#### Explore MCMC diagnostics

```{r fitModel_zinb_brm_MCMCdiagnostics, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Explore residual diagnostics

```{r fitModel_zinb_brm_diagnostics, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

**Conclusions:**

- no evidence of dispersion issues
- residual plot looks good
- no evidence of zero-inflation


#### Summarise model

```{r fitModel_zinb_brm_summary, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

::: {width="100%"}

```{r fitModel_zinb_brm_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
load(file='../data/mod.brm.RData')
emmeans(mod.brm, ~fYEAR, type='response') %>%
    as.data.frame %>%
    ggplot(aes(y=prob, x=as.numeric(fYEAR))) +
    geom_ribbon(aes(ymin=lower.HPD, ymax=upper.HPD), alpha=0.3) +
    geom_line()
```

:::

## Biomass {.tabset .tabset-faded}

### Delta method (INLA) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

I gave up on the following model.  It had been running for 30 minutes...

```{r fitModel_tweedie_INLA, results='markdown', eval=FALSE, fig.width=8, fig.height=5, cache=FALSE}
```

:::


### Delta method (INLA) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

```{r fitModel_detla_INLA, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=FALSE}
```

:::

::: {width="100%"}

```{r fitModel_delta_INLA_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

### Tweedie method (mgcv) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

```{r fitModel_tweedie_mgcv, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Explore diagnostics

```{r fitModel_tweedie_mgcv_diagnostics, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```


#### Summarise model

```{r fitModel_tweedie_mgcv_summary, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

::: {width="100%"}

```{r fitModel_tweedie_mgcv_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::



### Tweedie (glmmTMB) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

```{r fitModel_tweedie_glmmTMB, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Explore diagnostics

```{r fitModel_tweedie_glmmTMB_diagnostics, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Summarise model

```{r fitModel_tweedie_glmmTMB_summary, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

::: {width="100%"}

```{r fitModel_tweedie_glmmTMB_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

### ziGamma (glmmTMB) {.tabset .tabset-pills}

::: {width="100%"}

#### Fit model

Note, this takes a while

```{r fitModel_zigamma_glmmTMB, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Explore diagnostics

```{r fitModel_zigamma_glmmTMB_diagnostics, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

#### Summarise model

```{r fitModel_zigamma_glmmTMB_summary, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::

::: {width="100%"}

```{r fitModel_zigamma_glmmTMB_fig, results='markdown', eval=TRUE, fig.width=8, fig.height=5, cache=TRUE}
```

:::
